Beyond clines: lineages and haplotype blocks in hybrid zones

### 2016-04-25 drafts/review-responses-april-2016.tex

% \documentclass{article}
% \usepackage{fullpage}
% \input{review-response-commands}
% 
% \begin{document}

%%%%%%
%%
%%  Don't reorder the reviewer points; that'll mess up the automatic referencing!
%%
%%%%%

\begin{minipage}[b]{2.5in}
  Resubmission Cover Letter \\
  {\it Molecular Ecology}
\end{minipage}
\hfill
\begin{minipage}[b]{2.5in}
    Lewis-Sigler institute for integrative genomics\\
    Princeton University\\
    Yaniv's address\\
    \emph{and} Molecular and Computational Biology\\
    University of Southern California \\
  \today
\end{minipage}

@@

\begin{point}{}
``are have''
\end{point}

\reply
Fixed.

\begin{point}{\revref}
``reality'' means ``simulation'' here??
\end{point}

\reply
Well, we hope to mean ``reality'', but perhaps ``verify that theory matches reality'' is too strong; we've adjusted the wording to say ``\emph{realistic, individual-based models.}''.
%\alisa{I think we do mean ``reality" here, but alternate opinions welcome}
%\plr{I changed to ``realistic, individual-based models''.}

@@

\begin{point}{}
 Some appendices might be transferred to supplementary material -- the derivations are fairly standard.
\end{point}

\reply
\alisa{Our decision to keep the derivations in appendices was with consideration for the readership of Molecular Ecology. 
Thus, while some results are fairly standard, they might not be familiar to our readership.  
One of our concerns is that the math would be largely overlooked if moved into the supplement (which is not automatically printed with the main document) and this could create a barrier to understanding among the readers.}

@@

\begin{point}{}
    A serious limitation of the test is that it only detects selection against heterozygotes. The claim that this would not affect results, on l.453-456:
    \textit{``Although our assumed scenario of selection against heterozygotes at a single locus is uncommon in nature, we believe that this is unlikely to drastically influence our findings, as previous studies have demonstrated that underdominant loci share similar properties with more realistic models''}
    is unconvincing.
\end{point}

\reply
\alisa{We agree with the comments of the AE on the matter, and have added some additional text to the introduction that hopefully convinces that this is ok.}

\begin{point}{}
    From a theoretical perspective, the article makes modest but valid contributions, and  Equation 4 and 5 are potentially interesting extensions to equation (1). The practical benefit of using these equations rather than forward simulations appears limited.
\end{point}

\reply
\alisa{This is a point we have considered ourselves. We present these equations as a way of building our model up to address haplotype lengths. Both analytical expressions and simulations are presented so that they each support the conclusions of the other.} \yb{However, we also believe that analytical results can allow for a deeper understanding than simulation output. 
To aid our readers in developing this understanding, we have gone through considerable pains to map terms in our equations to the biological processes of migration, selection and recombination. 
We believe that our efforts will help build a better mechanistic understanding in our readers in addition to making them aware of our results.}

@@

\begin{point}{}
    While reading the manuscript, I found it difficult to distinguish the most important observations from more tangential observations. The manuscript reads like a long list of observations, and would benefit from more focus on the main findings and more structure.
\end{point}

\reply
\yb{We agree that a long set of theoretical results can be tedious. We have therefore reshaped the introduction to highlight our major question ``How does selection against incompatibilities in a hybrid zone influence the length distribution of ancestry tracts'' and result ``'It shortens them''.}

@@

\begin{point}{}
Since the recombination-based terms in Eqs 4 and 5 do not depend on the spatially explicit part of the model or on the selection term, I would be surprised if they had not been derived somewhere.
\end{point}

\reply
\yb{We agree that this representation has likely been derived many times. We turned to Hartl and Clark and found a similar derivation. We therefore signpost that this is a standard result by pointing readers to that text.}

@@

\begin{point}{Figure 3:}
What are the replicate lines for the same $r$? Why re the lines for $r = 0$ all slightly displaced to the right? Also, it would appear that the color scheme in associated Figure S2 (bottom) is mismatched with the legend, shifted to the red part of the spectrum. After fixing this, I think that the Figure S2 should be in the main text instead of Fig. 3.
\end{point}

\reply
\alisa{We believe that the displacement is likely due to drift. (We could do a repeated sim to show this?)}
\plr{I'm ok with switching S2 and 3; that seems like a good idea.  About the horizontal shift: I think that when I noticed this we figured out that it was because the center of the cline wasn't being plotted quite at zero for some reason?  If I recall correctly, we thought maybe it was drift but it wasn't.}
\plr{Alisa to clarify figure legend.}
\plr{peter or alisa fix S2: this is in \url{Plot_frequency_clines.R}}

\begin{point}{\revref}
Total population size is irrelevant when predicting the effects of genetic drift (besides, some simulations only have 50 demes across, not 100). In a one-dimensional habitat, one uses $N \sigma \sqrt{s}$ to judge whether the effect of genetic drift is likely to be negligible.
\end{point}

\reply
\yb{Thanks for spotting this. Because we do nothing quantitative with respect to drift here, we simply state the our observed wobble of the center of the tension zone is consistent with previous theory. However, more broadly throughout the manuscript we are not sure about  relying on $N \sigma \sqrt{s}$ as we are concerned with loci linked to the selected site, and we are not aware of results extending this rule to an unlinked locus.}
%\alisa{We are concerned here with linked loci, so the $N\sigma\sqrt{s}$ rule of thumb does not necessarily apply here? Also, within demes, population size is much smaller (500?)}
%\plr{I agree: they are probably thinking of selected loci, not linked ones. We're talking about global drift not local drift. Yaniv to investigate.}

### 2016-04-25 drafts/hybrid_zone.tex
yb did his best to deal with reviews

\paragraph{Linked clines}
We can use this diffusion model for lineages to find expected clines in ancestry,
i.e., the expected proportion of individuals who inherit from species $A$,
as a function of space, time, and position on the genome.
Precisely, we need the probability that 
an allele sampled $t$ generations after secondary contact at location $x$,
at recombination distance $r$ to a selected allele of type $z$,
is inherited from an individual of ancestry $A$,
where $z$ can be either $A$ or $B$.
We denote this probability $q_z(x,t,r)$.
In the notation above,
\begin{align}
    q_z(x,t,r) = \P^x \{Z_t = A \given X_0 = x, \; Z_0 = z\} .
\end{align}
(Recall that $Z$ depends implicitly on $r$.) \revpoint{1}{7}
The description above implies that $q_z$ solves the following Kolmogorov backward equation:
\begin{align}
    \begin{aligned}  \label{eqn:q_pde}
    \frac{d}{dt} q_A(x,t,r) 
            &= \sigma^2 \grad \log(p(x,t)) \cdot \grad q_A(x,t,r) 
                + \frac{\sigma^2}{2} \Delta q_A(x,t,r) 
            \\ &\qquad {} + 
                r (1-p(x,t))(q_B(x,t,r)-q_A(x,t,r))  \\
    \frac{d}{dt} q_B(x,t,r) &= \sigma^2 \grad \log(1-p(x,t)) \cdot \grad q_B(x,t,r) 
            + \frac{\sigma^2}{2} \Delta q_B(x,t,r)
            \\ &\qquad {} + 
            r p(x,t) (q_A(x,t,r)-q_B(x,t,r))  ,
    \end{aligned} 
\end{align}
with boundary conditions $q_A(x,0,r)=1$ and $q_B(x,0,r)=0$, 
and where $\grad$ is the gradient, and $\Delta$ is the Laplacian.
The three terms in these equations come from Brownian movement of a linages (i.e. the smoothing action of local dispersal),
the tendency of a lineage to inherit from regions where its type is more common (i.e. the net flux induced by reduced fitness in the hybrid zone),
and recombination between selected backgrounds, respectively. \alisa{I think the first and second terms may have been switched??}
%\alisa{this is fwd-in-time notation. should we change it to bwd-in-time?}
%\plr{This equation should be forwards in time, since it's the Kolmogorov backward equation
%applied to a backwards-time process!}

@@

As expected, clines in simulations form and begin to flatten, 
with ancestry frequencies at the selected site maintaining a stable shape,
and tightly linked sites flattening more slowly than distant sites.
This is seen, for instance, in Figures \ref{alleleFreq_tau100},
\ref{alleleFreq_tau100_weaker_s} and \ref{alleleFreq_tau1000}. 
Clines in ancestry frequency matched expected values found from deterministic theory
up to stochasticity due to genetic drift, which is more pronounced at lower population densities.
Figure \ref{sfig:alleleFreq_tau100_comparison} shows that expected clines computed from equation \eqref{eqn:q_pde}
match simulated clines in Figure \ref{alleleFreq_tau100} quite well.
Figures \ref{sfig:condAlleleFreq_tau80_comparison}, \ref{sfig:condAlleleFreq_tau320_comparison}, and \ref{sfig:condAlleleFreq_tau1280_comparison}
show this comparison with a lower population density than in Figure \ref{alleleFreq_tau100},
and separating linked clines based on the identity of the linked selected allele.
The agreement is good over hundreds of generations, but by $T=1280$ generations,
stochasticity has substantially moved allele frequencies range-wide
\citep[which is unsurprising, as the total population size is only 5,000 diploids:][]{Polechova2011}.
\revpoint{2}{6}

>>>>

\paragraph{Linked clines}
We can use this diffusion model for lineages to find expected clines in ancestry,
i.e., the expected proportion of individuals who inherit from species $A$,
as a function of space, time, and position on the genome.
Precisely, we need the probability that 
an allele sampled $t$ generations after secondary contact at location $x$,
at recombination distance $r$ to a selected allele of type $z$,
is inherited from an individual of ancestry $A$,
where $z$ can be either $A$ or $B$.
We denote this probability $q_z(x,t,r)$.
In the notation above,
\begin{align}
    q_z(x,t,r) = \P^x \{Z_t = A \given X_0 = x, \; Z_0 = z\} .
\end{align}
(Recall that $Z$ depends implicitly on $r$.) \revpoint{1}{7}
The description above implies that $q_z$ solves the following Kolmogorov backward equation:
\begin{align}
    \begin{aligned}  \label{eqn:q_pde}
    \frac{d}{dt} q_A(x,t,r) 
            &= \sigma^2 \grad \log(p(x,t)) \cdot \grad q_A(x,t,r) 
                + \frac{\sigma^2}{2} \Delta q_A(x,t,r) 
            \\ &\qquad {} + 
                r (1-p(x,t))(q_B(x,t,r)-q_A(x,t,r))  \\
    \frac{d}{dt} q_B(x,t,r) &= \sigma^2 \grad \log(1-p(x,t)) \cdot \grad q_B(x,t,r) 
            + \frac{\sigma^2}{2} \Delta q_B(x,t,r)
            \\ &\qquad {} + 
            r p(x,t) (q_A(x,t,r)-q_B(x,t,r))  ,
    \end{aligned} 
\end{align}
with boundary conditions $q_A(x,0,r)=1$ and $q_B(x,0,r)=0$, 
and where $\grad$ is the gradient, and $\Delta$ is the Laplacian.
The three terms in these equations come from Brownian movement of a linages (i.e. the smoothing action of local dispersal),
the tendency of a lineage to inherit from regions where its type is more common (i.e. the net flux induced by reduced fitness in the hybrid zone),
and recombination between selected backgrounds, respectively. \alisa{I think the first and second terms may have been switched??}
\yb{We note that because recombination occurs within a deme, the recombination terms in equation \eqref{eqn:q_pde} are quite standard  \citep{HartlClark}}
%\alisa{this is fwd-in-time notation. should we change it to bwd-in-time?}
%\plr{This equation should be forwards in time, since it's the Kolmogorov backward equation
%applied to a backwards-time process!}

@@

As expected, clines in simulations form and begin to flatten, 
with ancestry frequencies at the selected site maintaining a stable shape,
and tightly linked sites flattening more slowly than distant sites.
This is seen, for instance, in Figures \ref{alleleFreq_tau100},
\ref{alleleFreq_tau100_weaker_s} and \ref{alleleFreq_tau1000}. 
Clines in ancestry frequency matched expected values found from deterministic theory
up to stochasticity due to genetic drift, which is more pronounced at lower population densities.
Figure \ref{sfig:alleleFreq_tau100_comparison} shows that expected clines computed from equation \eqref{eqn:q_pde}
match simulated clines in Figure \ref{alleleFreq_tau100} quite well.
Figures \ref{sfig:condAlleleFreq_tau80_comparison}, \ref{sfig:condAlleleFreq_tau320_comparison}, and \ref{sfig:condAlleleFreq_tau1280_comparison}
show this comparison with a lower population density than in Figure \ref{alleleFreq_tau100},
and separating linked clines based on the identity of the linked selected allele.
The agreement is good over hundreds of generations, but by $T=1280$ generations,
stochasticity has substantially moved allele frequencies range-wide, \yb{which is in accordance with previous theory showing that tension zones wobble at a rate proportional $\sqrt{T}$ \citep{Barton1979a}. }
%\citep[which is unsurprising, as the total population size is only 5,000 diploids:][]{Polechova2011}.
\revpoint{2}{6}
